---
title: "coursework3"
author: "2010386"
date: "26/03/2021"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## INTRODUCTION

The data set consists of 2384 observations and 47 variables of police incidence which were obtained from January 2016 to January 2017.

This analysis would investigate police activities in Texas in the United State.It would explore the factors that leads to arrest,the crime rates and determine if there is an evidence of racial disparity in police activities .The following areas are going to be explored. 

* What is the monthly incident rate?
* Distribution of arrest by subject race. 
* Are traffic stop racially motivated?

 



```{r data ,echo=FALSE ,message =FALSE,include=FALSE}
#load the data and convert empty data to na 
mydata <- read.csv("../Assignment/37-00049_UOF-P_2016_prepped.csv",na.strings = c(""))
#remove the first row 
mydata <- mydata[-1,]
#convert empty Null to na 
mydata [mydata == "NULL"] <- NA
#view the data structure
str(mydata)
#change date format to date 
#tell r my date format first
mydata$INCIDENT_DATE <- as.Date(mydata$INCIDENT_DATE, format =  "%m/%d/%y")
#correct the date 
#get incident month in words by extracting from date
library(tidyverse)
mydata$INCIDENT_MONTH <- format(mydata$INCIDENT_DATE,"%m")
```

### Monthly Incident Rate

The data below shows a time trend of monthly incident count.The highest incident is observed in the month of march at a count of 264 and the second highest was in February with an incident count of 254 .The lowest incident was observed in the month of December at a count of 100.

```{r  crime rate,message =FALSE,echo=FALSE}
#crime montly distribution
#library(tidyverse)
mydata$INCIDENT_MONTH <- format(mydata$INCIDENT_DATE,"%m")
month_G1 <- mydata %>%
  drop_na(INCIDENT_MONTH) %>%
  group_by(INCIDENT_MONTH)%>%
   summarise(Cout= n()) 
# #plot monthly incident to see if there is a pattern 

monthly_crime <- month_G1 %>% ggplot(aes(x=INCIDENT_MONTH,y=Cout))
z <- monthly_crime+ geom_line(stat="identity",color ="purple",group =1) + 
  geom_point(size = 3, color = "steelblue") +
labs(title= "Monthly Incident Count" ,y ="Number of Incidents" , x= "Months")
library(plotly)
ggplotly(z)

```

The plot below shows the monthly incident count per race distribution.It can be seen in the first time trend that the incident count among the black race follows the overall trend with the highest incident count in March at 162 and lowest in December at 60.The second highest time trend was observed to be the Hispanic race with the highest month in January with 75 incidents , followed by the second highest month in march and the lowest incident count in December .Lastly, the third time trend which shows the white race incident count .It can be seen in the white time trend the highest peak observed was in April at 50 incident and similarly the second highest at October. The other three races were not plotted because they were negligible  compared to the incident count of white ,Hispanic and black.

```{r  crime rate per race ,message =FALSE,echo=FALSE}
#crime montly distribution
#library(tidyverse)
mydata$INCIDENT_MONTH <- format(mydata$INCIDENT_DATE,"%m")
month_race <- mydata %>%
  group_by(INCIDENT_MONTH,SUBJECT_RACE)%>%
   summarise(Cout= n())%>% drop_na()
# #plot monthly incident to see if there is a pattern 

monthly_race_incid <- month_race  %>%ggplot(aes(x=INCIDENT_MONTH,y=Cout))+ geom_line(data= filter(month_race,SUBJECT_RACE=="Black"),stat="identity",group =1,size=1,aes(color="purple"))+geom_line(data= filter(month_race,SUBJECT_RACE=="White"),stat="identity",group =1,size=1,aes(color="coral"))+
geom_line(data= filter(month_race,SUBJECT_RACE=="Hispanic"),stat="identity",group =1,size=1,aes(color="blue"))
#incident for other races are relatively small 
#make r recognise color through guide and map legend
monthly_race_incid +scale_color_identity(name = "Subject Race",
                          breaks = c("purple", "coral", "blue"),
                          labels = c("Black", "White", "Hispanic"),
                          guide = "legend")+
labs(title= "Monthly Incident Count" ,y ="Number", x= "Months" )



```

### Porportion of Subject Race Incidents that were Arrested and not Arrested

The bar graph below shows the portion of different subject race arrest and it was plotted with a categorical variable which indicated whether they were arrested or not .The y axis consists of the proportion of each race which was calculated as the count of a particular race arrested divided by the total number of incidence in the race.The graph shows a similar proportion of arrest in Black and Hispanic which is about 86% when encountered with a police incident.It was also seen that 88% of white who had a police incident were arrested. Lastly, Asian and American Indian were all arrested when involved in an incident.This leads to another question to which race is likely to get arrested.
```{r race arrest porportion during an incident,message =FALSE,warning=FALSE,fig.align='center',echo=FALSE}

library(scales)

library(plotly)


 sub_race_count <- table(factor(mydata$SUBJECT_WAS_ARRESTED),factor(mydata$SUBJECT_RACE))
 
 #find percentage of columns(subject race) and round up by 2 dp
sub_race <- round(100*prop.table(sub_race_count,2),2)
sub_race <-as.data.frame(sub_race)
sub_race <- sub_race  %>% rename(SUBJECT_WAS_ARRESTED =Var1,
                                 SUBJECT_RACE=Var2)
sub_race$label <- paste0(sub_race$label, "\n",
                         round(sub_race$Freq,2), "%")
# 
sub_race_count_plot <- sub_race %>% ggplot(aes(x=SUBJECT_RACE, y= Freq ,fill = SUBJECT_WAS_ARRESTED ))+ geom_bar(stat = "identity",position ="stack")+ 
 #geom_text(label = paste(Freq,"%"))+
 labs(title= "Percentage of Subject Race Incidents by arrest" ,y ="Percentage Incidents" , x= "Subject Race" , fill ="subject arrested")
  

ggplotly(sub_race_count_plot)
#plot the graph

```


The bar plot below shows the percentage proportion of arrest between different subject race. The highest percentage was observed from the Black race which came at about 56% , this was followed by almost a tie between Hispanic and white which are at 22% and 20% respectively.The last three races notably had low number of arrest compared to the top three races in the graph.
```{r race likely to have an incident,message =FALSE,warning=FALSE,fig.align='center',echo=FALSE}


 
 #find percentage of columns(subject race) and round up by 2 dp
sub_race_arrest <- mydata %>% filter(SUBJECT_WAS_ARRESTED =="Yes") %>% group_by(SUBJECT_RACE)%>% summarise(number = n()) %>% 
  drop_na()%>%
  arrange(desc(number))%>%
  mutate(percent = 100*number/sum(number))
# 
sub_race_arrest_plot <- sub_race_arrest %>% ggplot(aes(reorder(x=SUBJECT_RACE,-percent), y= percent ))+ geom_bar(stat = "identity",fill= "yellow",color="blue")+ 
 #geom_text(label = paste(Freq,"%"))+
 labs(title= "Percentage of Arrested Subject by Race" ,y ="Percentage Incidents" , x= "Subject Race")
  

ggplotly(sub_race_arrest_plot)
#plot the graph

```

### Incident Reason

The Incident reason was explored , the graph below shows the count of the top five incident reason .It can be seen from the graph that arrest made for the highest occurrence of a police incident at about 1157 , this was followed by service call at 673 and the last two were nearly similar for a police involvement which was traffic stop and crime in progress at an incident of93 and 82 respectively.The next visualization is to explore the race distribution among certain incident reason that can potentially show racial disparity.

```{r incident reason count,message =FALSE,warning=FALSE,echo=FALSE  }
# cateogry data ,sum wouldnt work
#is there racial bias in Traffic stop
incid_rsn <- mydata %>% group_by(INCIDENT_REASON) %>% 
  count() %>%
  arrange(desc(n))



incid_rsn_plot <- incid_rsn %>% head(5) %>% 
  ggplot(aes(reorder(x=INCIDENT_REASON,n),y= n, fill=n))+
  geom_bar(stat = "identity", color = "black") +  
scale_fill_gradient(low="orange", high="blue")
incid_rsn_plot + labs(title = "Top Five Incident Reason", x= "Incident Reason",y="Count", fill= "Count")

```

The pie chart below explores police incident reason for traffic stop .The figure shows the different subject race distribution .It can be observed that Blacks accounts for the highest percentage of traffic stop at about 52%, this is followed by Hispanic at 30.8% and the lowest observed was the Asian race at 1.1%.The other race and American Indian had no traffic stop incident according to this data.

```{r traffic stop pie plot,message =FALSE,warning=FALSE,echo=FALSE  }
# cateogry data ,sum wouldnt work
#is there racial bias in Traffic stop
library(scales) 
vehicle_stop <- mydata %>% 
  filter(INCIDENT_REASON == "Traffic Stop" ) %>%
  group_by(SUBJECT_RACE)%>%
  drop_na(SUBJECT_RACE)%>%
  summarise(number =n())%>%
  arrange(desc(SUBJECT_RACE)) %>% #calc prop
    mutate(prop = round(number * 100 / sum(number), 1),
           lab.ypos = cumsum(prop) - 0.5*prop)
#put label % line by line
vehicle_stop$label <- paste0(vehicle_stop$label, "\n",
                         round(vehicle_stop$prop,1), "%")

vehicle_stop_plot <- vehicle_stop %>% ggplot(aes(x=2,y=prop,fill= SUBJECT_RACE))+
  geom_bar(width =1, stat = "identity", color = "black") +  
coord_polar("y", start = 0,direction = 1)+
  geom_text(aes(y =lab.ypos ,label = label), color = "white") + xlim(0.5, 2.5) +
  theme_void()
vehicle_stop_plot +scale_fill_manual(values =c("plum2", "salmon1","gold2","maroon1"))
```


### Officer Year in the Force

The graph below shows an histogram of officer years in the force that are involved in an incident .The distribution can be seen to be right skewed and the most prominent peak at 1 to 3 years in the force ,this means officers in the first three years in the force are more likely to have an incident.

```{r ofcer histo by gender ,message =FALSE,warning=FALSE,echo=FALSE  }

#count number of officier years in the force
mydata$OFFICER_YEARS_ON_FORCE_cont <- as.numeric(mydata$OFFICER_YEARS_ON_FORCE)
library(plotly)
officer_yr_incid <- mydata %>% ggplot(aes(x=OFFICER_YEARS_ON_FORCE_cont)) +geom_histogram(fill= "bisque3",color="black",binwidth = 2)+
  scale_x_continuous(breaks = seq(0,36,2), limits = c(0,36))+
  theme_bw() +labs(title="Officer years in the force",y="Density",x="years")
ggplotly(officer_yr_incid) 

```

The next graph explored was the box plot for the officer years in the force versus the officer race .The first quartile (q1) this accounts for the 25th percentile of the distribution and third quartile (q3) accounts for 75th percentile of the distribution .This shows that 25% of the data for white officers fall about 1 year similarly for other races apart for American Indian at 4 years.The median was roughly the same among white ,other and Hispanic officer races.The highest average observed from all races was from the American Indian at 16 years.Outliers were observed in white officer years from 21years to 36 years possibly an error in the data collection or different method.Outliers are also seen in among Black and Hispanic officer.

```{r ofcer yrs box plot ,message =FALSE,warning=FALSE,echo=FALSE  }
mydata$OFFICER_YEARS_ON_FORCE_cont <- as.numeric(mydata$OFFICER_YEARS_ON_FORCE)
officer_yr_incid_box <- mydata %>% ggplot(aes(x=OFFICER_RACE,y=OFFICER_YEARS_ON_FORCE_cont)) +
  geom_boxplot(notch=TRUE,fill= "cyan4",alpha =0.7)+
  theme_bw() 
ggplotly(officer_yr_incid_box) 

```

### First Type of Force used

The graph below illustrates the count of the top ten first type of force used by police officer during an incident .The highest number of force used was verbal command at 818, the second highest observed was weapon display at person which accounts for a count of about 329 .The least occurring were taser display at person and take down body at about 68 count.The next thing to explore is the racial distribution among the type of force used to determine if there is racial disparity.

```{r subj weapon disp ,message =FALSE,warning=FALSE,echo=FALSE  }
#count number of first type of force used by officiers in the force
first_officer_force <- mydata %>% 
  count(TYPE_OF_FORCE_USED1)%>% 
  arrange(desc(n)) %>% 
  drop_na()

first_officer_force_plot <- first_officer_force %>% 
  head(10)%>% 
  ggplot(aes(reorder(x=TYPE_OF_FORCE_USED1,n), y= n))+ 
  geom_bar(stat="identity", color= "white" , fill="salmon1")+ 
  labs(title ="The top ten type of force used", x= "First Type of force used", y= "count")+ coord_flip() 

first_officer_force_plot + theme_minimal()


```

The bar graph below shows the percentage race distribution of weapon display display at subject.It can be seen that Black's account for 58.4% of weapon display during an incident , the least observed that were among American Indian, Asian and Other race . 

```{r officer weapon disp ,message =FALSE,warning=FALSE,echo=FALSE  }
#TYPE OF FORCE USED ON SUBJECT RACE
force_on_sub_race <- mydata %>% filter(TYPE_OF_FORCE_USED1 =="Weapon display at Person" )%>% group_by(SUBJECT_RACE) %>%
summarise(number =n())%>%mutate(prop = round(number * 100 / sum(number), 1))%>%
drop_na()


#label
force_on_sub_race$percentage <- paste0(force_on_sub_race$percentage, "\n",
                         round(force_on_sub_race$prop,1), "%")

sub_type1_weap_drw <- force_on_sub_race %>%
  ggplot(aes(x=reorder(SUBJECT_RACE,number),y=percentage))+ geom_bar(stat = "identity", fill = "lightblue",  color = "black")
 sub_type1_weap_drw + geom_text(aes(label=percentage))+
labs(x = "Subject Race",y = "Percent", title  = "Weapon drawn by Police at Subject Race distribution")
 text <-mydata %>%filter(TYPE_OF_FORCE_USED1 =="Weapon display at Person" )%>%
   select(SUBJECT_RACE,SUBJECT_WAS_ARRESTED)
 #text
 #table(text$SUBJECT_WAS_ARRESTED,text$SUBJECT_RACE)
```

The bar graph shows the distribution of different officer race that displayed weapon at the first the course of action.The white officer have the highest percentage at 62% and the least observed were from American Indian and other race at 0.6%.

```{r 1st type of force ,message =FALSE,warning=FALSE,echo=FALSE  }
 #which  OFficer race is likely to draw weapon
 
 force_by_offic_race <- mydata %>% filter(TYPE_OF_FORCE_USED1 =="Weapon display at Person" )%>% group_by(OFFICER_RACE) %>%
summarise(number =n())%>%mutate(prop = round(number * 100 / sum(number), 1))%>%
drop_na()
#label a percentage column
force_by_offic_race$percentage <- paste0(force_by_offic_race$percentage, "\n",
                         round(force_by_offic_race$prop,1), "%")

force_by_offic_race_plot <- force_by_offic_race %>%
  ggplot(aes(x=reorder(OFFICER_RACE,number),y=percentage))+ geom_bar(stat = "identity",  color = "black",fill="coral")
 force_by_offic_race_plot + geom_text(aes(label=percentage))+
labs(x = "Officer Race",y = "Percent", title  = "Weapon drawn by Police distribution")
 
```

### Incident Injury 

The figures below shows the count of injury observed from both officer and subject race distribution.The first figure shows the officer injury by race .It can be seen that white officers had the highest count of injury at about 160 and the lowest was injury observed was from Asian officer which was notably small. 

The second figure shows the injury count of different race subject.The highest injury was observed from Black subjects at about 343 ,the second highest observed was from white subject and Asian subject had no injury.

```{r officer injury/sub injury ,message =FALSE,warning=FALSE,echo=FALSE  }
officer_hs <- mydata %>% count(OFFICER_INJURY)%>% 
  arrange(desc(n)) %>% 
  drop_na()

#race explanatory , injury response
sub_hs <- mydata %>%group_by(SUBJECT_RACE)%>%
  count(SUBJECT_INJURY)%>% 
  arrange(desc(n)) %>% 
  drop_na()
officier_injury <-ggplot(mydata,aes(x=OFFICER_RACE, fill = OFFICER_INJURY)) + geom_bar(position ="dodge")+
scale_y_continuous(breaks = seq(0,1500,250), limits = c(0,1500))
officier_injury+ labs(title = "Officer Race Injury Count", y= "Count", x="Officer Race")

subject_injury <- mydata %>% drop_na(SUBJECT_RACE)%>%
  arrange(desc(SUBJECT_RACE))%>%
  ggplot(aes(x=SUBJECT_RACE,fill= SUBJECT_INJURY,y=..count..))+geom_bar(position ="dodge")+
scale_y_continuous(breaks = seq(0,1000,250), limits = c(0,1000))
subject_injury+labs(title = "Subject Race Injury Count", y= "Count", x="Subject Race")

sub_inj <- ftable( mydata$SUBJECT_RACE~ mydata$SUBJECT_INJURY)
sub_inj_count <- round(prop.table(sub_inj,2),2)
sub_inj_count <- as.data.frame(sub_inj_count)


```


```{r crime ,message =FALSE,warning=FALSE,echo=FALSE,include=FALSE  }
sub_crime <- mydata %>% count(SUBJECT_OFFENSE)%>% 
  arrange(desc(n)) %>% 
  drop_na()%>% mutate(Total=cumsum(n))
pattern <- "^Drug Possession /|Drug Possession "
s <- c("Drug Possession  - Misdemeanor","Drug Possession  - Felony")
str_detect(s, "^Drug")

mydata$grp_crime <- mydata$SUBJECT_OFFENSE %>%
  str_replace("^Assa\\D+","Assualt") %>%
  str_replace("^Warr\\D+","Warrant")%>%
  str_replace("^Resisting Arrest\\D+|Resisting Arrest|Evading Arrest\\D+","Resisting Arrest") %>%
  
  str_replace("^Drug Possession\\D+|Drug Possession|Drug Possession[.-]|Man/Del/Dangerous Drug\\D+","Drug Possession ")%>%
  str_replace("^Traffic Violation\\D+|Traffic Violation","Traffic Violation") %>%
  str_replace("^Warrant/|Warrant","Warrant")%>%
  str_replace("^Burglary of a Vehicle\\D+|Burglary of a Vehicle|Burglary\\D+|Burglary","Burglary") %>%
  str_replace("^Public Intoxication\\D+|Public Intoxication ","Public Intoxication")%>%
  str_replace("^Other Misdemeanor Arrest\\D+|Other Misdemeanor Arrest","Misdemeanor Arrest") %>%
  str_replace("^Other Felony Arrest\\D+|Other Felony Arrest","Felony")%>%
  str_replace("^DWI\\D+|DWI|DWLI|DUI|DUI\\D+","Driving under influence") %>%
  str_replace("^Theft\\D+|Theft","Theft")%>%
  str_replace("^Fail to ID\\D+","Fail to ID")%>%
  str_replace("^Resisting Search\\D+|Resisting Search","Resisting Search") %>%
  str_replace("^Robbery\\D+","Robbery")%>%
  str_replace("^APOWW\\D+","APOWW")%>%
  str_replace("^Discharge Firearm\\D+","Discharge Firearm")%>%
  str_replace("^Unauthorize Use Vehicle\\D+","Unauthorize Use Vehicle") %>%
  str_replace("^Disorderly Conduct\\D+","Disorderly Conduct")%>%
  str_replace("^Crim Trespas\\D+","Tresspass")

#remove no arrest


```

### Incident Division Distribution 

The bar graph below shows the percentage proportion of incident rate among different division .The y axis shows the division and x-axis shows the proportion.The highest percentage observed across the divisions was observed in south central with 87.3% accounting to black subject incident and it also had the lowest white subject arrest at 4.5% and Hispanic arrest at 8.1% across all the seven division.Northeast and southeast had similar proportion for black subject incident at 65.6%.In the southwest there is an almost balanced incident observed between black and hispanic subjects coming at 46.7% and 43.9% accordingly, the same balance was also observed in north central among white and black subjects .The black subject had high police incident across five district with Northwest accounting for the lowest at 28.6%

```{r division incident,message =FALSE,warning=FALSE,echo=FALSE }
library(leaflet)
mydata$Latitude <- as.numeric(mydata$LOCATION_LATITUDE)
mydata$Longitude <- as.numeric(mydata$LOCATION_LONGITUDE)
#group incident by division
division_incid <-mydata %>% 
  group_by(DIVISION,SUBJECT_RACE)%>%
  summarise(number=n())%>%drop_na()%>%
  mutate(prop=100*round(number/sum(number),3))

division_incid_plot <-division_incid %>% ggplot(aes(x=DIVISION,y=prop,fill=SUBJECT_RACE))+
  geom_bar(stat="identity",color="grey") +coord_flip()+theme_bw()+
  scale_fill_manual(values=c("yellow","blue","cyan2","slategray1","pink","plum1"))
ggplotly(division_incid_plot)




```

### Crime Map
The map below shows the distribution of crime.The first crime stated in an incident was grouped under one category to enable visualisation.It shows the subject offense from resisting arrest to APOWW and more.

```{r crime map ,message =FALSE,warning=FALSE,echo=FALSE  }


library(htmltools)
library(leaflet)
mydata$Latitude <- as.numeric(mydata$LOCATION_LATITUDE)
mydata$Longitude <- as.numeric(mydata$LOCATION_LONGITUDE)


mymap_crime <- leaflet(mydata) %>%
 addTiles() %>%
   addCircleMarkers(data = mydata[mydata$grp_crime=="Drug Possession ",], group = "Drugs",col="red",label = ~htmlEscape("Drugs"))%>%
  addCircleMarkers(data = mydata[mydata$grp_crime=="Theft",], group = "Theft",color="yellow",label = ~htmlEscape("Theft"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="Assualt",], group = "Assualt",color="#fee090",label = ~htmlEscape("Assualt"))%>%
  addCircleMarkers(data = mydata[mydata$grp_crime=="Warrant",], group = "Warrant",color="purple",label = ~htmlEscape("Warrant"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="Burglary",], group = "Burglary",color="orange",label = ~htmlEscape("Burglary"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="Tresspass",], group = "Tresspass",color="peach",label = ~htmlEscape("Tresspass"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="APOWW",], group = "APOWW",color="light blue",label = ~htmlEscape("APOWW"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="Robbery",], group = "Robbery",color="#4575b4",label = ~htmlEscape("Robbery"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="Resisting Arrest",], group = "Resisting Arrest",color="#74add1",label = ~htmlEscape("Resisting Arrest"))


mymap_crime%>%  addLayersControl(
    baseGroups = c("OSM (default)", "Toner Lite"),
    #define layer group into overlay object
    overlayGroups = c("Theft","Drugs", "Assualt","Burglary","Tresspass","APOWW", "Warrant", "Resisting Arrest","Robbery"),
    #layer options to show one layer at a time
    options = layersControlOptions(collapsed = TRUE))


```

The map below shows the distribution of incident reason in the city of Dallas .The traffic stop distribution seems to be spread with a few concentration in certain parts of the city.It also shows a high distribution of arrest in the city .

```{r incident reason map ,message =FALSE,warning=FALSE,echo=FALSE  }
incid_rsn <- mydata %>% group_by(INCIDENT_REASON) %>% 
  count() %>%
  arrange(desc(n))

 

incident_map <-leaflet(mydata) %>%
  addTiles() %>% 
addCircleMarkers(lat=subset(mydata,INCIDENT_REASON=='Traffic Stop')$Latitude, lng=subset(mydata, INCIDENT_REASON=='Traffic Stop')$Longitude ,color = "red",label = ~htmlEscape('Traffic Stop'),group='Traffic Stop') %>%
  addCircleMarkers(lat=subset(mydata,INCIDENT_REASON=='Arrest')$Latitude, lng=subset(mydata, INCIDENT_REASON=='Arrest')$Longitude ,color = "blue",label = ~htmlEscape('Arrest'),group='Arrest')%>%
  addCircleMarkers(lat=subset(mydata, INCIDENT_REASON=='Call for Cover')$Latitude, lng=subset(mydata, INCIDENT_REASON=='Call for Cover')$Longitude, color = "yellow",opacity = 0.2,label = ~htmlEscape('Call for Cover'),group= 'Call for Cover')%>%
  addCircleMarkers(lat=subset(mydata, INCIDENT_REASON=='Crime in Progress')$Latitude, lng=subset(mydata, INCIDENT_REASON=='Crime in Progress')$Longitude, color = "purple",opacity = 0.2 ,label = ~htmlEscape('Crime in Progress'),group= 'Crime in Progress')%>%
  addCircleMarkers(lat=subset(mydata,INCIDENT_REASON=='Service Call')$Latitude, lng=subset(mydata, INCIDENT_REASON=='Service Call')$Longitude ,color = "pink",label = ~htmlEscape('Service Call'),group= 'Service Call')

incident_map %>%addLayersControl(
    baseGroups = c("OSM (default)", "Toner Lite"),
    #define layer group into overlay object
    overlayGroups = c('Traffic Stop','Arrest', 'Call for Cover',"Crime in Progress","Service Call"),
    #layer options to show one layer at a time
    options = layersControlOptions(collapsed = TRUE))

```

## Conclusion

The data was explored mainly using these variables such as first type of forced used, injury,incident reason,subject offense,subject race and officer race,Incident date and subject arrest.The data shows that there seems to be a disproportion of arrest among black subject ,this indicates a need for exploration into the population of black subjects among different district to explain this observation. It was also seen that black subjects were likely to get stopped at traffic stop this seems to indicate a bias ,the same observation was found among Hispanic subjects.In addition,due to the high number of black subject involved in police incident, they were observed to have the highest injury during an incident .Lastly, the first type of force used that was used as weapon display, it was observed that black subjects were more likely to be shown a weapon during an incident .More data needs to obtained from at least 5 years coupled with population data , subject income, subject job, subject age to reach a definite conclusion.